# Good Practice in Data Analysis

Tidy data is a standard way of mapping the meaning of a dataset to its structure. A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types. In **tidy data**:

1.  Each variable forms a column.

2.  Each observation forms a row.

3.  Each type of observational unit forms a table.

**Messy data** is any other arrangement of the data.

Tidy data makes it easy for an analyst or a computer to extract needed variables because it provides a standard way of structuring a dataset. 

Tidy data is particularly well suited for vectorised programming languages like R, because the layout ensures that values of different variables from the same observation are always paired.

## Tidying messy datasets

Real datasets can, and often do, violate the three precepts of tidy data in almost every way imaginable.This section describes the five most common problems with messy datasets, along with their remedies:

-   Column headers are values, not variable names.

-   Multiple variables are stored in one column.

-   Variables are stored in both rows and columns.

-   Multiple types of observational units are stored in the same table.

-   A single observational unit is stored in multiple tables.

### Column headers are values, not variable names

A common type of messy dataset is tabular data designed for presentation, where variables form both the rows and columns, and column headers are values, not variable names.

The following code shows a subset of a typical dataset of this form. This dataset explores the relationship between income and religion in the US.

```{r}
pew <- tbl_df(read.csv("https://raw.githubusercontent.com/tidyverse/tidyr/master/vignettes/pew.csv", stringsAsFactors = FALSE, 
                       check.names = FALSE))
pew
```
When gathering variables, we need to provide the name of the new key-value columns to create. The first argument, is the name of the key column, which is the name of the variable defined by the values of the column headings. In this case, it's `income`. The second argument is the name of the value column, `frequency`. The third argument defines the columns to gather, here, every column except religion.

```{r}
pew %>%
  gather(income, frequency, -religion)
```

This form is tidy because each column represents a variable and each row represents an observation, in this case a demographic unit corresponding to a combination of `religion` and `income`.

## Multiple variables stored in one column

After gathering columns, the key column is sometimes a combination of multiple underlying variable names. This happens in the `tb` (tuberculosis) dataset, shown below. This dataset comes from the World Health Organisation, and records the counts of confirmed tuberculosis cases by `country`, `year`, and demographic group. The demographic groups are broken down by `sex` (m, f) and `age` (0-14, 15-25, 25-34, 35-44, 45-54, 55-64, unknown).

```{r}
tb <- tbl_df(read.csv("https://raw.githubusercontent.com/tidyverse/tidyr/master/vignettes/tb.csv", stringsAsFactors = FALSE))
tb
```

First we gather up the non-variable columns:

```{r}
tb2 <- tb %>% 
  gather(demo, n, -iso2, -year, na.rm = TRUE)
tb2
```

Column headers in this format are often separated by a non-alphanumeric character (e.g. `.`, `-`, `_`, `:`), or have a fixed width format, like in this dataset. `separate()` makes it easy to split a compound variables into individual variables. You can either pass it a regular expression to split on (the default is to split on non-alphanumeric columns), or a vector of character positions. In this case we want to split after the first character:

```{r}
tb3 <- tb2 %>% 
  separate(demo, c("sex", "age"), 1)
tb3
```

Storing the values in this form resolves a problem in the original data. We want to compare rates, not counts, which means we need to know the population. In the original format, there is no easy way to add a population variable. It has to be stored in a separate table, which makes it hard to correctly match populations to counts. In tidy form, adding variables for population and rate is easy because they're just additional columns.

## Variables are stored in both rows and columns

The most complicated form of messy data occurs when variables are stored in both rows and columns. The code below loads daily weather data from the Global Historical Climatology Network for one weather station (MX17004) in Mexico for five months in 2010. 

```{r}
weather <- tbl_df(read.csv("https://raw.githubusercontent.com/tidyverse/tidyr/master/vignettes/weather.csv", stringsAsFactors = FALSE))
weather
```

It has variables in individual columns (`id`, `year`, `month`), spread across columns (`day`, d1-d31) and across rows (`tmin`, `tmax`) (minimum and maximum temperature). Months with fewer than 31 days have structural missing values for the last day(s) of the month.

To tidy this dataset we first gather the day columns:

```{r}
weather2 <- weather %>%
  gather(day, value, d1:d31, na.rm = TRUE)
weather2
```

For presentation, I've dropped the missing values, making them implicit rather than explicit. This is ok because we know how many days are in each month and can easily reconstruct the explicit missing values.

We'll also do a little cleaning:

```{r}
weather3 <- weather2 %>% 
  mutate(day = extract_numeric(day)) %>%
  select(id, year, month, day, element, value) %>%
  arrange(id, year, month, day)
weather3
```

This dataset is mostly tidy, but the `element` column is not a variable; it stores the names of variables. (Not shown in this example are the other meteorological variables `prcp` (precipitation) and `snow` (snowfall)). Fixing this requires the spread operation. This performs the inverse of gathering by spreading the `element` and `value` columns back out into the columns:

```{r}
weather3 %>% spread(element, value)
```

This form is tidy: there's one variable in each column, and each row represents one day.

