---
title: "Statistical Testing"
author: "Rajarshi Guha"
date: "January 5, 2015"
output:
  html_document:
    toc: true
    theme: readable
    number_sections: true
---

As you might expect R provides extensive support for statistical testing. In this section we'll focus on a few examples, though we won't go into the details of the underlying theory.

# Hypothesis testing

## Testing for normality

Many hypothesis tests as well as modeling methods such as least squares regression assume that your data is normally distributed.  So it's a good idea to test this assumption before further calculations. Lets look at the potency values for the actives from a MIPE4 screen and use some visual methods to check for normality.
```{r fig.width=10}
dat <- read.csv('/ncats/prod/common/R-Workshop/mipe4-qhts.csv',
                header=TRUE,as.is=TRUE,comment='')
actives <- subset(dat, CCLASS2 %in% c(-1.1, -1.2))
potencies <- actives$LAC50

par(mfrow=c(1,2))
# Not very normal
qqnorm(potencies)
qqline(potencies)

# Plot the density distrubution
plot(density(potencies))
```

We can also use the [Shapiro Wilk](http://en.wikipedia.org/wiki/Shapiro%E2%80%93Wilk_test) test
```{r}
shapiro.test(potencies)

# Just to be sure, what does the test say for really normal data?
normal.values <- rnorm(100)
shapiro.test(normal.values)
```

There are other tests for normality ([Kolmogorov Smirnov](http://en.wikipedia.org/wiki/Kolmogorov%E2%80%93Smirnov_test), [Anderson-Darling](http://en.wikipedia.org/wiki/Anderson%E2%80%93Darling_test), [Cramer von Mises](http://en.wikipedia.org/wiki/Cram%C3%A9r%E2%80%93von_Mises_criterion), etc.). You can usually find the corresponding R functions via a Google search or consider the [nortest](http://cran.r-project.org/web/packages/nortest/nortest.pdf) package for a one stop shop. Also see [Ghasemi and Zahediasi](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3693611/) for a more scholarly description.

However, [normality tests don’t do what you think they do](http://www.r-bloggers.com/normality-tests-don%E2%80%99t-do-what-you-think-they-do/)

## Testing for differences in means

The [t-test](http://en.wikipedia.org/wiki/Student%27s_t-test) is used in a variety of screening scenarios. A t-test can be used for 

* A one-sample location test of whether the mean of a population has a value specified in a null hypothesis.
* A two-sample location test of the null hypothesis such that the means of two populations are equal (assuming equal variances), also termed a "independent samples" t-test
* A test of the null hypothesis that the difference between two responses measured on the same statistical unit has a mean value of zero, also termed a  "paired" or "repeated measures" t-test

For example, given a MIPE screen run in two cell lines - a wild type and a mutant. We could ask

* For a given compound, does its potency differ between the two lines, in a statistically significant manner?
* Is there a difference in the potency of the actives between the two cell lines?

The former can't really be tested for unless we have replicate data. The second can be addressed using the t-test (though if the data are not normal it's better to use the [Mann Whitney U (aka Wilcoxon rank-sum)](http://en.wikipedia.org/wiki/Mann%E2%80%93Whitney_U_test) test).  

Another useful task is to see whether there is a difference in the potencies between the two cell lines, but on a target-wise basis. An approach to do this for MIPE screens is [available](http://carnot.ncats.nih.gov:3838/tdiff/)

## Testing for enrichment

In many cases it's useful to test whether certain items in a subset are overrepresented compared to the entire dataset. A more concrete example is to ask whether the any of the targets corresponding to the actives in a qHTS screen are overrepresented compared to the distribution of the targets in the entire MIPE collection. We can answer this question by using the [hypergeometric test](http://en.wikipedia.org/wiki/Hypergeometric_distribution) (a.k.a., [Fishers exact test](http://en.wikipedia.org/wiki/Fisher%27s_exact_test)).

Here's a function to do this in the general cases, where you have a sample of symbols taken from the population of symbols
```{r}
enrichment <- function(moas, pop.moas, meth) {
  umoas <- sort(unique(moas))
  if (!all(umoas %in% unique(pop.moas))) 
    stop("All symbols being tested must occur in the background population")
  pvals <- sapply(umoas, function(amoa) {
    sampleSize <- length(moas)
    hitInSample <- length(which(moas == amoa))
    hitInPop <- length(which(pop.moas == amoa))
    failInPop <- length(pop.moas) - hitInPop
    fisher.test(matrix(c(hitInSample-1,
                         hitInPop-hitInSample,
                         sampleSize-hitInSample,
                         failInPop-sampleSize+hitInSample),2,2), alt='greater')$p.value
  })
  data.frame(moa=umoas, p=pvals, padj=p.adjust(pvals, meth))
}
```
For example, lets say we have a population of symbols and a random sample of them
```{r}
pop <- c(rep('A', 70), rep('B', 30), rep('C', 40), rep('D', 40))
sub <- c(rep('A', 20), rep('B', 15), rep('C', 5), rep('D', 3))
enrichment(sub, pop, 'BH')
```

This is a contrived example so lets go back and see what targets are enriched in the set of actives from a MIPE4 screen. As before we load in our data from the CSV and then identify a subset of actives. As our population (or background) we consider all non-`NA` MIPE4 targets
```{r}
dat <- read.csv('/ncats/prod/common/R-Workshop/mipe4-qhts.csv',
                header=TRUE,as.is=TRUE,comment='')
actives <- subset(dat, CCLASS2 %in% c(-1.1, -1.2))
actives.targets <- actives$target
all.targets <- dat$target

# we remove NA's
actives.targets <- na.omit(actives.targets)
all.targets <- na.omit(all.targets)

enr <- enrichment(actives.targets, all.targets, 'BH')

## order by adjusted p-value and select targets enriched at alpha = 0.05
enr <- enr[order(enr$padj),]
subset(enr, padj < 0.05)
```

### GO term enrichments
Another common enrichment calculation involves the use of [GO](http://geneontology.org/) terms. For this type of analysis you wouldn't use the above function since the hierarchical nature of GO terms requires some adjustments. However, R provides a handy method for this analysis via the [GOstats](http://www.bioconductor.org/packages/release/bioc/html/GOstats.html) package. Rather than using gene symbols, we'll need Entrez Gene IDs, which we can also obtain using R
```{r cache=TRUE, message=FALSE, warning=FALSE}
library(GOstats)
library(org.Hs.eg.db)

# Some target symbols do not map to gene ids. Example is FGFR which actually should
# be FGFR1 or FGFR2 or ...
actives.geneid <- mget(actives.targets, envir=org.Hs.egSYMBOL2EG, ifnotfound=NA)
uni.geneid <- mget(all.targets, envir=org.Hs.egSYMBOL2EG, ifnotfound=NA)
params <- new("GOHyperGParams", 
              geneIds = actives.geneid, 
              universeGeneIds=uni.geneid, 
              ontology = "BP",
              pvalueCutoff = 0.05, conditional=F, 
              testDirection = "over", annotation="org.Hs.eg.db")
r <-  hyperGTest(params)
gosum <- summary(r)
gosum[1:10,]
```

# Adjusting p-values

The outcome of most hypothesis tests are p-values. Leaving aside a discussion of the abuse of p-values, you should be aware of multiple hypothesis testing and the corresponding adjustment of the p-values. There are many ways to perform these adjustments - [Bonferroni](http://en.wikipedia.org/wiki/Bonferroni_correction), [Benjamini-Hochberg](http://en.wikipedia.org/wiki/False_discovery_rate), etc. 

Given a set of p-values from a set of hypothesis tests, we can obtain the adjusted p-values using `p.adjust` and specifying the adjustment method. My personal preference is 'BH'. Note that the `p.adjust` function doesn't support Storey's [q value](http://genomics.princeton.edu/storeylab/qvalue/) method. For that consider using the [qvalue](http://www.bioconductor.org/packages/release/bioc/html/qvalue.html) package.

# Analysis of variance

Another way to look at differences in means is to perform an Analysis of Variance (ANOVA). While this may be overkill for a simple comparison of, say, potencies, between two conditions, it becomes very useful when considering multiple effects (say, effect of target or target and dose of sensitizer and so on). 

Lets consider the MIPE4 screen and see whether the target of the compound has an effect on its potency. We consider compounds with good curve classes and annotated with a target. Our first questio is, *does the target have an effect on potency?*
```{r}
dat <- read.csv('/ncats/prod/common/R-Workshop/mipe4-qhts.csv',
                header=TRUE,as.is=TRUE,comment='')
good <- subset(dat, CCLASS2 %in% c(-1.1, -1.2) & !is.na(target))
model <- aov(LAC50 ~ target, good)
summary(model)
```

The summary of the ANOVA indicates that that target has a statistically significant effect on the potency. Next we're interested in seeing whether there are differences in potency between targets. We can do this using a post-hoc test
```{r}
posthoc <- TukeyHSD(model)
posthoc <- data.frame(posthoc$target)
subset(posthoc, p.adj < 0.05) # only need the significant cases
```

Interestingly, this suggests that potency differences between [TUBB](http://en.wikipedia.org/wiki/TUBB) and a few other targets are statistically significant. This may or may not be interesting!

We can get more complex. Consider two MIPE screens - one run on a cell line with [CD47](http://en.wikipedia.org/wiki/CD47) and one with it KO'd. Does the KO status and/or the target affect the potency? Does the combination of KO and target affect the potency?
```{r}
dat <- read.csv('/ncats/prod/common/R-Workshop/mipe4-4t1b.csv',
                header=TRUE,as.is=TRUE,comment='')
datws <- read.csv('/ncats/prod/common/R-Workshop/mipe4-4t1b+CD47.csv',
                header=TRUE,as.is=TRUE,comment='')
good <- subset(dat, CCLASS2 %in% c(-1.1, -1.2) & !is.na(target))
goodws <- subset(datws, CCLASS2 %in% c(-1.1, -1.2) & !is.na(target))

tmp <- rbind( data.frame(cond='4T1B', good[, c('LAC50', 'target')]),
              data.frame(cond='4T1B+CD47', goodws[, c('LAC50', 'target')]))
model <- aov(LAC50 ~ target * cond, tmp)
summary(model)

posthoc <- TukeyHSD(model)
posthoc$cond
posthoc.target <- data.frame(posthoc$target)
subset(posthoc.target, p.adj < 0.01) # only need the significant cases
```

The analysis indicates that while target does affect potency, the potency is not affect by KO status (or by the interaction of KO status and target)
