---
title: "(Very Basic) R Programming"
author: "Rajarshi Guha"
date: "January 5, 2015"
output:
  html_document:
    toc: true
    theme: readable
    number_sections: true

---

# Why program in R

For many tasks you can use R just like SPSS or Stata - and it will likely be a bit inconvenient compared to those packages. For example you could read in CSV's files of qHTS data for two screens corresponding to two treatments and perform a t-test to see if there is a statistically significant difference in potency. 

Even slicing and dicing data (such as subsetting or transforming columns) can usually be performed in a few easy steps. However, in many cases, you need to perform repetitive operations - say perform some calculation on multiple datasets or combine a series of operations into a 'workflow' and so on.

For these scenarios, being able to construct loops and write functions will save you a lot of time and lead to elegant (and in many cases) more efficient code.

# Functions in R

This is a very brief intro to writing R functions - I'll be glossing over many details. See [here](http://nicercode.github.io/guides/functions/) for a more detailed introduction and [here](http://adv-r.had.co.nz/Functions.html) for a more advanced description of functions.

A function consists of 4  things

* The function name (which can be any valid R variable name)
* The function arguments (what it will accept - which could be nothing)
* The body of the function (the actual stuff that the function will do)
* The return value of the function

Consider a simple function that takes in a single argument and returns that argument plus 2
```{r}
f1 <- function(x) {
  return(x+2)
}
```

From the description (or by reading the code) its safe to assume that you send a *number* into the function. But there's nothing stopping you from sending in a string - if you do it'l fail.  
```{r collapse=TRUE, error=TRUE}
f1(2)
f1("Hello")
```

This is a good thing and a bad thing in R - without reading the documentation it's not obvious what type of data goes in to a function. But at the same time this allows you to write a function that can take any data type and deal with it appropriately.

Importantly, if the code inside a function is written appropriately, you can send in different data structures. Since addition works element-wise on a vector we can send in a single number or a vector of numbers. But lists don't support element-wise operations.
```{r collapse=TRUE, error=TRUE}
f1(c(1,2,3,4))
f1(list(1,2,3,4))
```

R functions can have arguments that have a default value. Lets change the above function so that we can specify what number to add to `x`
```{r collapse=TRUE}
f2 <- function(x, what.to.add = 2) {
  return(x + what.to.add)
}
f2(2) # Will use the default value of what.to.add
f2(2, 5)
f2(2, what.to.add = 5)
```

In these examples we've returned a single value. In many cases it can be useful to return multiple values (e.g., after fitting a dose response curve you want to return the 4 fit parameters). In this cases, use a list with named elements as the return value
```{r collapse=TRUE, error=TRUE}
f3 <- function(x, a) {
  r <- list(value=x+a, addend=a)
  return(r)
}
f3(2) # fails because a has no default value
f3(2, 10)
```

## Variable scope

When it comes to determining the value of a variable there is a path that R will use to search for its value. In the case of functions if a previously undefined variable appears R will look at the argument list for the function. Next it will look in the current work space. If you are not careful R will find the value some place where you do not expect it, and your function will return a value that is not correct, and no error will be given. **Be very careful about the names of variables especially when using functions.**

## Everything is a function

While `1+2` doesn't look like a function, the `+` operator is actually implemented as a function, so that you can write `"+"(1,2)` and it will return `r "+"(1,2)`. For now this is a curiosity but it does allow you to redefine any function (so `+` could be redefined to mean `-` - not a great idea!)

# Loops in R

The for loop in R works by operating over a sequence (which could be a vector or a list) of values
```{r}
for (i in c(1,2,3,4)) {
  print(i)
}
for (j in list(1,2,3,4)) {
  print(j+2)
}
```

As a simple example, given a vector we want to obtain the vector squared, element-wise
```{r}
v <- c(1,2,3,4)
sv <- numeric(length(v))
for (i in v) {
  i^2 # but how do we save the squared value?
}
for (i in 1:length(v)) {
  sv[i] <- v[i]^2
}
sv
```

## `apply` and friends

When using `for` loops, saving the result of each iteration usually involves either preparing a vector (or other data structure) to contain that result, or else keep track of a index variable ourselves. R provides a few other functions that return the result of operating on a vector or list as another vector or list or matrix. These functions include `apply`, `sapply`, `lapply`, `mapply`, and `tapply`. See this [Stackoverflow answer](http://stackoverflow.com/a/7141669) for a nice description of these functions. A quick summary:

* `lapply` is an apply which returns a list (acts on any vector or list and returns a list)
* `sapply` is a simple lapply (function defaults to returning a vector or matrix when possible)
* `rapply` is a recursive apply (for nested lists, i.e. lists within lists)
* `tapply` is a tagged apply (the tags identify the subsets)
* `apply` is generic: (applies a function to a matrix's rows or columns)

(These functions represent what is called [functional programming](http://en.wikipedia.org/wiki/Functional_programming). If you've never worked in this approach, an R-specific [introduction](http://www.bioinformaticszen.com/post/simple-functional-programming-in-r/) is useful.)

The `sapply` method operates on a vector or list and applies a function (provided by the user) to each element of the vector and returns these result of the function as another vector.
```{r}
v <- 1:10
sapply(v, function(x) x^2)
```

(Here we've used what is called an *anonymous function* - it has no name and is only used in the context of the `sapply` function. You'll see that it doesn't show up in your *Environment* panel)

When you provide the function to `sapply` you have to ensure that it accepts the appropriate data type. If it's written to accept numbers, then the above code will fail if `v` is a character vector.

The `apply` function is useful when operating on entire rows or columns of a matrix. Say we want the mean response (across all compounds) at each dose of a qHTS experiment. 
```{r}
dat <- read.csv('/ncats/prod/common/R-Workshop/mipe4-qhts.csv',
                header=TRUE,as.is=TRUE,comment='')
resps <- dat[, 12:22]
apply(resps, 2, mean)
```

Here the second argument, `2` indicates that the function (`mean`) will receive each column of the data.frame in turn. Importantly, the function passed to `apply` must support a single vector as input (since a row or a column of the matrix is a vector). Whether this is a numeric vector or a character vector is problem dependent - but you need to anticipate what type of data will be sent to the function. If we specified `1` instead, then `mean` would receive each *row* of the matrix (so you'd get the mean response across all doses, for each compound)
```{r}
mean.resp.by.cmpd <- apply(resps, 1, mean)
length(mean.resp.by.cmpd)
```

While the function that is applied must accept the vector as input, it can have other arguments. These can have default values; but if not, you must specify the value of the argument when calling `apply` (and this applies to all the apply family of functions)
```{r collapse=TRUE}
resps[1,1] <- NA
apply(resps, 2, mean) # we should ignore NA's
apply(resps, 2, mean, na.rm=TRUE)
```

### Iterating over subsets

While these functions iterate over every element (or along a dimension), a common task is to operate on subsets of the data. For example, what is the median AUC and median potency for each curve class in a qHTS experiment? `by` comes to the rescue
```{r}
r <- by(dat, dat$CCLASS2, function(x) c(median(x$auc), median(x$LAC50)))
r
do.call(rbind, as.list(r))
```

In this example, `by` iterates over subsets of `dat` where each subset corresponds to a unique value of `CCLASS2` and passes the subset (which is a `data.frame`) to the anonymous function. Since this function receives a `data.frame` we can operate on specific columns. The return value in this case is a special object of class `by` - we can convert this to a list using `as.list` and then transform the list to a matrix.

However, `tapply` can provide a useful, simpler alternative if you want to summarize a single variable across combinations of groups. For example, what is the mean potency conditioned on curve class?
```{r}
tapply(dat$LAC50, dat$CCLASS2, mean)
```

The nice thing about `tapply` is that you get a simple R data type (array or matrix) rather than a special object that requires extra handling. We could also get a curve class *and* target-wise summary:
```{r}
m <- tapply(dat$LAC50, list(dat$CCLASS2, dat$target), mean, na.rm=TRUE)
m[1:10, 1:5] # it's a big matrix, so we just show a part of it
```

So, if you want to operate on entire subsets of a `data.frame` consider using `by`. But if you want to operate on a single variable of a `data.frame` subsetted by one or more other variables, consider `tapply`.
