# Data Pipelines

Uptil this point we've been using base R - all of the material can be rerun without installing extra packages. However, it can be tedious and especially if you do a lot of transformations of your data it can be difficult to follow.

```{r}
library(tidyverse)
```

The `tidyverse` set of packages is a nice way to work with data in a way that makes it easier to understand whats going on and also adhere to best practices.

For example, though a `data.frame` is a base R data type, the `tibble` is a handy method to work with `data.frame`'s that offers a variety of conveniences
```{r, echo=TRUE}
tbl <- as_tibble(dat.clean)
tbl
```

The `dplyr` package provides a number of functions to operate on chunks of data by filtering, splitting and combining rows of a `data.frame`. Previously we used `subset`. The corresponding function in `dplyr` is `filter`
```{r}
tmp <- dplyr::filter(dat.clean, Vehicle.Color == 'Red')
head(tmp)
```

Before we go any further, we should exploit the new pipe operator that the tidyverse imports from the  `magrittr` package. This new syntax leads to code that is much easier to write and to read.
```{r}
dat.clean %>% head()
```

Using the pipe operator (`%>%`) we can "chain" various operations in easily understandable sequences. For example, lets take a subset of the columns, rename one of them, and filter the results, and report the number of days parked aggregated over ZIP codes

The old way:

```{r}
tmp <- dat.clean[, c('Vehicle.Make.Model', 'Vehicle.Color', 'ZIP.Code',
                     'How.Many.Days.Has.the.Vehicle.Been.Reported.as.Parked.',
                     'Police.District', 'Community.Area')]
names(tmp)[4] <- 'Days.Parked'
tmp <- subset(tmp, Days.Parked < 365)
by(tmp, tmp$ZIP.Code, function(x) mean(x$Days.Parked))
```

Using the pipe operator and `dplyr` functions we have 
```{r}
result <- dat.clean %>%
  dplyr::select(Vehicle.Make.Model, Vehicle.Color, ZIP.Code,
                Completion.Date,
                Police.District, Community.Area, 
                Days.Parked = How.Many.Days.Has.the.Vehicle.Been.Reported.as.Parked.) %>%
  dplyr::filter(Days.Parked < 365)
  
result %>% 
  group_by(ZIP.Code) %>%
  summarize(mean.days = mean(Days.Parked))
```

## Handling categorical data

We've already discussed categorical variables which are called factors in R, and how we can deal with them. The `tidyverse` provides the `forcats` packages which makes it a bit easier to deal with factors.

```{r}
library(forcats)
dat.clean %>% dplyr::count(Vehicle.Make.Model)
```

We could directly create a barchart of counts for each make/model
```{r fig.width=10}
dat.clean %>% ggplot(aes(Vehicle.Make.Model))+geom_bar()+
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

For now lets focus on the vehicle makes that occur more than a 500 times
```{r}
makes <- dat.clean %>% dplyr::count(Vehicle.Make.Model) %>% dplyr::filter(n>500)
```

And then just keep those rows of our dataset where the vehicle make occurs more than 200 times:
```{r}
dat.makes <- dat.clean %>% dplyr::inner_join(makes, by='Vehicle.Make.Model')
```
Now, if we plot the frequencies of each make, we get a more reasonable figure
```{r fig.width=10}
dat.makes %>% ggplot(aes(Vehicle.Make.Model))+geom_bar()+
  theme(axis.text.x = element_text(angle=45, hjust=1))
```
But it would be nice if we reordered the levels of the factor in order of increasing frequency. `fct_infreq` lets us do that for plotting
```{r fig.width=10}
dat.makes %>% 
  mutate(make = Vehicle.Make.Model %>% fct_infreq()) %>%
  ggplot(aes(make))+
  geom_bar()+theme(axis.text.x = element_text(angle=45, hjust=1))
```

After reordering factors, a useful task is to _recode_ factors. This allows us to rename a specific level or group multiple levels into a single one. If we look at the vehicle makes with more than 500 occurences we have the following list
```{r}
levels(dat.makes$Vehicle.Make.Model)
```
Lets, collapse all the `Jeep` makes into a single make called `Jeep`
```{r}
dat.makes <- dat.makes %>%
  mutate(Vehicle.Make.Model = fct_recode(Vehicle.Make.Model,
      'Jeep' = "Jeep - Year Greater 1988",
      'Jeep' = "Jeep (If Mfd. Prior To 1970)",
      'Jeep' = "Jeep/Cherokee"))
```
If we're collapsing a lot of levels, `fct_collapse` is an easier way to achieve that
```{r}
dat.makes <- dat.makes %>%
  mutate(Vehicle.Make.Model = fct_collapse(Vehicle.Make.Model,
      "Range Rover" = c("Range Rover of North America", "Range Rover Of North America"),
      'Honda' = c("Honda", "Honda/Accord", "Honda/Civic"),
      'Ford' = c("Ford", "Ford/Escort", "Ford/Taurus")))
```